---
title: "Class 12 HW: Population analysis"
author: "Lily Huynh (PID: A16929651)"
date: 2/13/2025
toc: true
format: pdf
---
## Section 4: Population Scale Analysis 

> Q13. Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. 

> Ans. A/A sample size: 108, median expression levels (A/A): 31.24847; A/G sample size: 233, median expression levels (A/G): 25.06486; G/G sample size: 121, median expression levels (G/G): 20.07363

```{r}
sample.table <- read.table("rs8067378_ENSG00000172057.6.txt", header=TRUE)

head(sample.table)
```

```{r}
summary(sample.table)
```

I will use the `table()` function to figure out the sample size of each genotype. 

```{r}
sample.size.table <- table(sample.table$geno)
sample.size.table
```

Then I will make sort the data into different tables based on the genotype. 

```{r}
library(dplyr)
genoAA <- sample.table %>%
  filter(geno == "A/A")
head(genoAA)

genoAG <- sample.table %>%
  filter(geno == "A/G")
head(genoAG)

genoGG <- sample.table %>%
  filter(geno == "G/G")
head(genoGG)
```

Finally, I will calculate the median for each genotype 

```{r}
median(genoAA$exp)
```

```{r}
median(genoAG$exp)
```

```{r}
median(genoGG$exp)
```

> Q14. Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

> Ans. The relative expression value of G/G is lower compared to A/A. Therefore, alleles with G will have a lower expression compared to alleles with A. Based on the boxplot, I can see that the SNP does effect the expression of ORMDL3 because when A/A is changed to A/G, the expression levels are lower. 

I created two different boxplots, using the function `boxplot()` and `ggplot()`.

```{r}
boxplot(genoAA$exp, genoAG$exp, genoGG$exp)
```

```{r}
library(ggplot2)
ggplot(sample.table) + 
  aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```

