---
title: "Comparative structure analysis"
author: "Lily Huynh (PID: A16929651)"
format: html
toc: true
---

Here we run through a complete "pipeline" of structure analysis that begins with a single sequence identifer and ends in a PCA analysis. 

```{r}
library(bio3d)
```

## Step 1. Get a sequence
Retrive a sequence for the protein we are interest in. We will take ADK "1ake_A"

```{r}
id <- "1ake_A"
aa <-get.seq(id)
aa

```

## Step 2. BLAST structure
Run a BLAST search of the PDB for all related sequences to our input `aa`

```{r}
blast <- blast.pdb(aa)
```

```{r}
hits <- plot(blast)
```

## Step 3. Download all structures
These are our "top hits" i.e. all the structures in the PDB database related to our input sequence.

```{r}
hits$pdb.id
```

```{r}
# Download related PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

## Step 4. Align and superpose 

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

```{r}
pdbs
```

## Step 5. PCA

Let's use our old friend PCA to make sense of these confusing, complicated structure relationships. 

```{r}
pc <- pca(pdbs)
plot(pc)
```

Let's make a trajectory (or movie) of the main confrontational changes captured by PC1. We will use the 'mktrj()` function for this...

```{r}
mktrj(pc, file="pca_result.pdb")
```

Back of the envelope comparison of the PDB size to UniProt

```{r}
uniprot <- 253206171
pdb <- 231029
pdb/uniprot *100
```




































